Live coding environment for a data exploration language

### 2019-02-04 wip-live-programming/paper.tex
Adding empirical evaluation section

As an example, the anlysis done by the Financial Times for an article on plastic waste
\cite{ftnotebooks,ftarticle} joins datasets from Eurostat, UN Comtrade and more, aggregates the data and
builds a visualization comparing waste flows in 2017 and 2018. Figure~\ref{fig:ft-uncomtrade} shows
an extract from one notebook from the data analysis. The code has a number of remarkable properties:
%
\begin{itemize}
\item There is no abstraction. The analysis uses lambda functions as arguments to library calls,
  but it does not define any top-level functions. Parameterization is done by defining a list of
  inputs and then having a for loop, or by defining a top-level variable and setting its value.
  In our example, \ident{material} is set to \str{plastics} and a comment lists other options.
  This lets the analyst easily see and check the results of intermediate steps of the analysis.

\vspace{0.5em}
\item The code relies on external libraries. Our example uses Pandas \cite{pandas},
  which provides opreations for data wrangling such as \ident{merge} to join datasets
  or \ident{drop\_duplicates} to delete rows with duplicate column values. Those libraries are
  external to the data analysis and may even be implemented in other languages -- many Python
  libraries rely on C++.

\vspace{0.5em}
\item The code is structured as a sequence of commands. Some commands define a variable, either by
  loading data, or by transforming data loaded previously. Even in Python, data is often treated
  as immutable. Other commands produce an output that is displayed in the notebook.

\vspace{0.5em}
\item There are many corner cases, such as the fact that the \ident{keep\_default\_na} parameter
  needs to be set to handle Namibia correctly. These are discovered interactively by writing and
  running a code snippet and examining the output, so providing a rapid feedback is essential.
\end{itemize}

@@

\begin{figure}
\includegraphics[scale=0.5]{drawing.png}
\caption{Zzz}
\label{fig:drawing}
\end{figure}

\subsubsection{Methodology}

image.load("shadow.png").greyScale().blur(5)

image.load("shadow.png").greyScale().blur(10)

let s = image.load("shadow.png").greyScale().blur(5)
s

let s = image.load("shadow.png").greyScale().blur(10)
s.combine(image.load("pope.png"), 80)

let r = 80
let s = image.load("shadow.png").greyScale().blur(10)
s.combine(image.load("pope.png"), r)

\subsubsection{Conclusions}

@@

>>>>

As an example, the anlysis done by the Financial Times for an article on plastic waste
\cite{ftnotebooks,ftarticle} joins datasets from Eurostat, UN Comtrade and more, aggregates the data and
builds a visualization comparing waste flows in 2017 and 2018. Figure~\ref{fig:ft-uncomtrade} shows
an extract from one notebook from the data analysis. The code has a number of remarkable properties:
%
\begin{itemize}[itemsep=3pt]
\item There is no abstraction. The analysis uses lambda functions as arguments to library calls,
  but it does not define any top-level functions. Parameterization is done by defining a list of
  inputs and then having a for loop, or by defining a top-level variable and setting its value.
  In our example, \ident{material} is set to \str{plastics} and a comment lists other options.
  This lets the analyst easily see and check the results of intermediate steps of the analysis.

\item The code relies on external libraries. Our example uses Pandas \cite{pandas},
  which provides opreations for data wrangling such as \ident{merge} to join datasets
  or \ident{drop\_duplicates} to delete rows with duplicate column values. Those libraries are
  external to the data analysis and may even be implemented in other languages -- many Python
  libraries rely on C++.

\item The code is structured as a sequence of commands. Some commands define a variable, either by
  loading data, or by transforming data loaded previously. Even in Python, data is often treated
  as immutable. Other commands produce an output that is displayed in the notebook.

\item There are many corner cases, such as the fact that the \ident{keep\_default\_na} parameter
  needs to be set to handle Namibia correctly. These are discovered interactively by writing and
  running a code snippet and examining the output, so providing a rapid feedback is essential.
\end{itemize}

@@

The key performance claim about our method of providing instantaneous feedback is that it is more
efficient than recomputing values for the whole program (or the current command) after every
keystroke. In the previous section, we formally proved that this is true and gave examples of
code edit operations that do not cause recomputation. In this section, we further support this
claim with an empirical evaluation. The purpose of this section is not to precisely evaluate
overheads of our implementation, but to compare how much recomputation different evaluation
strategies perform.

For the purpose of the evaluation, we use a simple image manipulation library that provides
operations for loading, greyscaling, bluring and combining images. We compare delays in
updating the preview for three different evaluation strategies, while performing the same
sequence of code edit operations. Using image processing as an example gives us a way to
visualize the reuse of previously computed values. As in a typical data exploration scenario,
the individual operations are relatively expensive compared to the overheads of building the
dependency graph.

\begin{figure}
\raggedright

\vspace{0.5em}\hspace{1.5em}{\small\sffamily (1) Enter the following code and then change parameter of blur from 4 to 8:}
\begin{equation*}
\begin{array}{l}
\ident{image}.\ident{load}(\str{shadow.png}).\ident{greyScale}().\ident{blur}(\num{4})
\end{array}
\end{equation*}

\vspace{0.5em}\hspace{1.5em}{\small\sffamily (2) Assign the result to a variable and start writing code for further operations:}
\begin{equation*}
\begin{array}{l}
\kvd{let}\,\ident{shadow}=\ident{image}.\ident{load}(\str{shadow.png}).\ident{greyScale}().\ident{blur}(\num{8})\\
\ident{shadow}.\ident{combine}
\end{array}
\end{equation*}

\vspace{0.5em}\hspace{1.5em}{\small\sffamily (3) Finish code to combine two images and then change parameter of combine from 20 to 80:}
\begin{equation*}
\begin{array}{l}
\kvd{let}\,\ident{shadow}=\ident{image}.\ident{load}(\str{shadow.png}).\ident{greyScale}().\ident{blur}(\num{8})\\
\ident{shadow}.\ident{combine}(\ident{image}.\ident{load}(\str{pope.png}), \num{20})
\end{array}
\end{equation*}

\vspace{0.5em}\hspace{1.5em}{\small\sffamily (4) Extract the parameter of combine to a let bound variable:}
\begin{equation*}
\begin{array}{l}
\kvd{let}\,\ident{ratio} = \num{80} \\
\kvd{let}\,\ident{shadow}=\ident{image}.\ident{load}(\str{shadow.png}).\ident{greyScale}().\ident{blur}(\num{8})\\
\ident{shadow}.\ident{combine}(\ident{image}.\ident{load}(\str{pope.png}), \ident{ratio})
\end{array}
\end{equation*}

\caption{Code edit operations that are used in the experimental evaluation}
\label{fig:image-edits}
\end{figure}

\begin{figure}[b]
\vspace{-0.5em}
\includegraphics[scale=0.425]{drawing.png}
\caption{Time required to recompute the results of a sample program after individual tokens of
  the program are added or modified for three different evaluations strategies.}
\label{fig:drawing}
\vspace{-0.5em}
\end{figure}

\subsubsection{Code edit operations and methodology}

To avoid clutter when visualizing the performance, we update the preview after complete tokens are
added (such as identifier, number, dot or left parenthesis) rather than after individual keystrokes.
Figure~\ref{fig:image-edits} shows the sequence of code edit operations that we use to measure
the delays in updating a live preview. We first enter an expression to load, greyscale and blur
an image (1) then introduce let binding (2) and add more operations (3). Finally, we extract
one of the parameters into a variable (4). Most of the operations are simply adding code, but
there are two cases where we modify existing code and change value of a parameter for \ident{blur}
and \ident{combine} immediately after (1) and (3), respectively.

We implement the algorithm described in Sections~\ref{sec:formal} and \ref{sec:previews} in a
simple web-based environment that allows the user to modify code and explicitly trigger
recomputation. It then measures time needed to recompute values for the whole program and
displays the resulting image. If the parsing fails, e.g.~because of an unclosed parenthesis,
previews are not updated and we record only the time taken by parsing attempt. We compare the
delays of three different evaluation strategies:

\begin{itemize}[itemsep=3pt]
\item \emph{Call-by-value.} Following the semantics in Section~\ref{sec:calculus-semantics},
  all sub-expressions are fully evaluated before an expression is evaluated. In this case, work may
  be done even if the evaluation results in an error. For example, we parse
  $\ident{image}.\ident{load}(\str{shadow.png}).\ident{blur}$ as member access with no arguments.
  The evaluation loads the image, but then fails because blur requires one argument.

\item \emph{Lazy.} To address the wastefulness of call-by-value strategy discussed above, we
  simulate lazy evaluation by implementing a version of the image processing library where
  operations build a delayed computation and only evaluate it when rendering an image. Using
  this strategy, failing computations do not perform unnecessary work.

\item \emph{Live.} Finally, the live strategy uses the algorithm described in
  Sections~\ref{sec:formal} and \ref{sec:previews}. The cache is empty at the beginning of the
  experiment and we update it after each token is added. This is the only strategy that does
  not start afresh after reparsing code.
\end{itemize}

\subsubsection{Efficiency evaluation}
The experimental environment is implemented in F\# and compiled to JavaScript using the Fable
compiler. We run the experiments in Firefox (version 64.0.2, 32bit) on Windows 10
(1809, 64bit) on a computer with Intel Core i7-7660U CPU and 16Gb RAM.

Figure~\ref{fig:drawing} shows times needed to recompute previews after individual tokens
of the program are added, deleted or modified, according to the script in
Figure~\ref{fig:image-edits}, resulting in 38 measurements for each of the 3
evaluation strategies. We mark a number of notable points in the chart:

\begin{enumerate}[(a)]
\item Loading image for the first time incurs small extra overhead in the live strategy.
\item Greyscaling using the live strategy does not need to re-load the image.
\item Accessing the blur member without arguments causes delay for call-by-value strategy.
\item When varying the parameter of blur, the live strategy reuses previously greyscaled image.
\item Introducing let binding does not cause recomputation when using live strategy.
\item As in (c), accessing combine member without valud argument only affects call-by-value.
\item Varying parameter of combine in live strategy does not recompute blur and is much faster.
\item Introducing let binding, again, causes full recomputation for lazy can call-by-value.
\end{enumerate}

\begin{figure}
\noindent
\includegraphics[scale=0.48]{hist-all.png}
\includegraphics[scale=0.48]{hist-slow.png}
\caption{Histogram showing the distribution of delays incurred when updating previews.
  We show a histogram computed from all delays (left) and only from delays larger
  than 15ms (right).}
\label{fig:drawing-hist}
\end{figure}

\noindent
A summarized view of the delays is provided in Figure~\ref{fig:drawing-hist}, which shows
histograms illustrating the distribution of delays for each of the three evaluation methods.
A large proportion of delays is very small (less than 15ms) because the parser used in our
experimental environment often fails (e.g.~for unclosed parentheses). The histogram
on the right summarizes only delays for edit operations where the delay for at least one of
the strategies was over 15ms. A more robust parser would be able to recover and the delays
for call-by-value and lazy strategies would be even more significant.

\subsubsection{Summary of results}
The purpose of our empirical is not to exactly assess the overhead of our implementation
of the live evaluation strategy algorithm. Instead, our goal is to
illustrate how often can previously evaluated results be reused and the impact this has when
writing code. The experiment presented in this section is small-scale, but it is sufficient
to illustrate the main point.

When recomputing results after every edit operation
using the \emph{call-by-value} strategy, the time needed to update the results grows
continually. The \emph{lazy} strategy removes the overhead for programs that fail to evaluate, but
keeps the same trend. Using our \emph{live} strategy, the computation can reuse values computed
previously and, as a result, expensive operations (such as (d) and (g) in Figure~\ref{fig:drawing})
are significantly faster, because they do not need to recompute operations done previously when
writing the code. As shown in Figure~\ref{fig:drawing-hist}, there are almost no very expensive
operations (taking over 1 second) in the \emph{live} strategy in contrast to several in the
other two strategies.

@@

http://gallery.thegamma.net/75/gender-pay-gap
http://gallery.thegamma.net/73/libraries
http://gallery.thegamma.net/47/unemployment-in-italy
http://gallery.thegamma.net/44/classification-of-the-functions-of-government-cofog
http://gallery.thegamma.net/37/top-10-government-departments-by-201516-expenditure
